---
title: "MA615-HW4"
author: "Haochen Li"
date: "2024-09-27"
output:
  pdf_document:
    latex_engine: xelatex
---
\maketitle

\begin{center}
\end{center}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(data.table)
library(lubridate)
library(dplyr)
library(tidyr)
library(readxl)
library(ggplot2)
library(pROC)

setwd('E:/Desktop/BU/2024 Fall/MA615/HW4')
data <- read.csv("E:\\Desktop\\BU\\2024 Fall\\MA615\\HW4\\final_data1.csv")
rain <- read.csv("E:\\Desktop\\BU\\2024 Fall\\MA615\\HW4\\Rainfall.csv")

```

# B: Your next exercise is to identify and deal with the null data in the dataset...

```{r}
# Replace extreme values-NA

data <- data %>%
  mutate(YYYY = ifelse(YYYY < 100, YYYY + 1900, YYYY))%>%
  mutate(across(where(is.numeric), ~ na_if(., 99))) %>%
  mutate(across(where(is.numeric), ~ na_if(., 999))) %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))

summary(data)

```

### Answer: Replacing 'extreme' data may not always be suitable. For instance, in cases where 999 could represent a valid extreme measurement rather than missing data, replacing it with NA might obscure significant insights.

```{r}
# Convert- Plot
data$date <- as.Date(paste(data$YYYY, data$MM, data$DD, sep="-"))

# Count
na_counts <- sapply(data, function(x) sum(is.na(x)))
na_df <- data.frame(Variable = names(na_counts), NA_Counts = na_counts)

# Plot
ggplot(na_df, aes(x = NA_Counts, y = reorder(Variable, NA_Counts))) +
  geom_bar(stat = "identity", fill = "Orange") +
  geom_text(aes(label = NA_Counts), 
            position = position_stack(vjust = 0.7),
            size = 3.5, color = "Black") +
  labs(title = "NA Distribution Across Variables", x = "Number of NA", y = "Variables") +
  theme_minimal()
```

### The NA distribution reveals that variables like TIDE, WD, BAR, and VIS have the most missing data, with each having over 400,000 entries marked as NA. On the other hand, variables such as WTMP and PRES.BAR show much fewer missing values, with as low as 13,000.

# C: Can you use the Buoy data to see the effects of climate change?...

### Answer: Yes. The analysis of maximum water, air, and dew point temperatures <br> over the decades shows a consistent upward trend across all seasons, reinforcing the <br> impact of climate change. Across Spring, Summer, Fall, and Winter, the data reveals that the highest temperatures for water (WTMP), air (ATMP), and dew point (DEWP) have steadily increased from the 1985-1994 period to the 2015-2023 period. This rise in both sea surface and atmospheric temperatures reflects broader environmental changes, including the potential for more extreme weather events and disruptions in ecosystems. The evidence strongly supports the conclusion that climate change is driving significant increases in both water and atmospheric temperatures.

```{r}
# Convert- date
data <- data %>%
  mutate(date = as.Date(date))

# Group by month
data <- data %>%
  mutate(year_month = floor_date(date, "month"))

# Group by decade
data <- data %>%
  mutate(decade = case_when(
    YYYY >= 1985 & YYYY <= 1994 ~ "1985-1994",
    YYYY >= 1995 & YYYY <= 2004 ~ "1995-2004",
    YYYY >= 2005 & YYYY <= 2014 ~ "2005-2014",
    YYYY >= 2015 & YYYY <= 2023 ~ "2015-2023"
  ))

# Remove rows where ATMP, WTMP, and DEWP are all NA
data <- data %>%
  filter(!(is.na(ATMP) & is.na(WTMP) & is.na(DEWP)))

# Data 4 plot
Max_temp_MM <- data %>%
  group_by(decade, year_month) %>%
  summarise(
    max_ATMP = ifelse(all(is.na(ATMP)), NA, max(ATMP, na.rm = TRUE)),
    max_WTMP = ifelse(all(is.na(WTMP)), NA, max(WTMP, na.rm = TRUE)),
    max_DEWP = ifelse(all(is.na(DEWP)), NA, max(DEWP, na.rm = TRUE)),
    .groups = 'drop'
  )

# Group by Season
Max_temp_MM <- Max_temp_MM %>%
  mutate(
    month = month(year_month),
    season = case_when(
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall",
      month %in% c(12, 1, 2) ~ "Winter"
    ),
    season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter"))
  )


# Boxplot of Max Water Temperature by Season and Decade
ggplot(Max_temp_MM, aes(x = season, y = max_WTMP, fill = decade)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Max Water Temperature by Season and Decade",
    x = "Season",
    y = "Max Water Temperature (°C)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +  # Apply color palette
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Boxplot of Max Air temperature by Season and Decade
ggplot(Max_temp_MM, aes(x = season, y = max_ATMP, fill = decade)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Max Air temperature by Season and Decade",
    x = "Season",
    y = "Max Air temperature (°C)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +  # Apply color palette
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Boxplot of Max Dewpoint temperature by Season and Decade
ggplot(Max_temp_MM, aes(x = season, y = max_DEWP, fill = decade)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Max Dewpoint temperature by Season and Decade",
    x = "Season",
    y = "Max Dewpoint temperature (°C)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +  # Apply color palette
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# D: As part of this Homework, you have been given data for rainfall in Boston from 1985 to the end of 2013.

```{r}
rain$DATE <- ymd_hms(rain$DATE)
rain <- rain %>% arrange(DATE)

summary_stats <- rain %>%
  summarise(
    mean_rainfall = mean(HPCP, na.rm = TRUE),
    median_rainfall = median(HPCP, na.rm = TRUE),
    total_count = n(),
    max_rainfall = max(HPCP, na.rm = TRUE),
    min_rainfall = min(HPCP, na.rm = TRUE),
    sd_rainfall = sd(HPCP, na.rm = TRUE)
  )

# Display the summary table
print(summary_stats)
ggplot(rain, aes(x = DATE, y = HPCP)) +
  geom_line(color = "blue") +
  labs(title = "Rainfall Over Time in Boston", x = "Date", y = "Rainfall (inches)")

# Create target variable
rain <- rain %>%
  mutate(HPCP_next_hour = lead(HPCP, n = 1),# shift HPCP for next hour prediction
         Rain_next_hour = ifelse(HPCP_next_hour > 0, 1, 0)) %>%
  drop_na(HPCP_next_hour)

# Add time features
rain <- rain %>%
  mutate(hour = hour(DATE),
         day_of_week = wday(DATE),
         month = month(DATE))

# Data for logistic regression
rainfall_data_clean <- rain %>% select(hour, day_of_week, month, Rain_next_hour)
X <- rainfall_data_clean[, c('hour', 'day_of_week', 'month')]
y <- rainfall_data_clean$Rain_next_hour

# Split data into train and test sets
set.seed(123)
train_index <- sample(1:nrow(X), 0.7 * nrow(X))
X_train <- X[train_index,]
X_test <- X[-train_index,]
y_train <- y[train_index]
y_test <- y[-train_index]

# Fit the logistic regression model
train_data <- data.frame(X_train, Rain_next_hour = y_train)

# Fit the logistic regression model all data
logistic_model <- glm(Rain_next_hour ~ hour + day_of_week + month, 
                      data = train_data, family = binomial)

# Make predictions
P_y <- predict(logistic_model, newdata = X_test, type = "response")
y_pred <- ifelse(P_y > 0.5, 1, 0)

# Evaluate model performance
accuracy <- mean(y_pred == y_test)
Roc <- roc(y_test, P_y)
auc_value <- auc(Roc)

# Plot ROC Curve
ggplot(data = data.frame(fpr = Roc$specificities, tpr = Roc$sensitivities), aes(x = fpr, y = tpr)) +
  geom_line() +
  geom_abline(linetype = "dashed", color = "grey") +
  labs(title = sprintf("ROC Curve (AUC = %.2f)", auc_value), 
       x = "False Positive Rate", y = "True Positive Rate")
```

### The graph shows periodic rainfall spikes in Boston throughout 2020, with rainfall reaching up to 2 inches during certain events, followed by periods of no rain.

### The model has weak predictive power, with the ROC curve near the diagonal and an AUC slightly above 0.5. Time-based features alone aren't sufficient for accurate rainfall prediction. Adding more relevant features like temperature, pressure, and humidity would likely improve performance.